Survey of Cache Oblivious B-trees

### 2011-12-12 paper.tex
finished restructuring co string b-trees section for now

\usepackage{paralist}
\usepackage{graphicx}

@@

Two experiments are performed, in which a regular B-tree implementation tuned
to various block sizes is compared to a static COSB-tree implementation. The
experiment involved measuring the amount of time required to perform: 440,000
and 450,000 inserts of random values, a range query of all values, and 1000
random searches. The results of the experiment show that COSB-trees
outperformed B-trees most of the time for random searches, range queries and
random inserts. However, some point between 440,000 and 450,000 inserts the
COSB-tree had to rebalance the entire tree, causing it to fall behind in this
measurement. However, Bender et al. argue that the experiment was biased
towards the B-tree implementation, since the trees are "young", and thus still
well organized. The COSB-tree implementation is also compared to Berkeley DB
**cite berkeley db**, and achieves performance comparable to Berkeley DB with
tuned parameters. The results seem to show that there is a window in which
large and small-block size B-trees perform poorly, and that COSB-trees may be
a reasonable alternative to improve performance.

@@

The static COSB-tree consists of 2 layers: a centroid tree at the top, and an
array of the keys stored in lexicographic order on the bottom. The leaves of
the centroid tree contain pointers to keys in the array. The centroid tree is
structured such that a search through the centroid tree identifies the local
area of the desired key in the array. Then a sequential scan is performed
across the array to find the key.

The Dynamic COSB-tree consists of 3 major components: a centroid tree at the
top; a packed memory array (PMA) structure of hash data for the keys known as
the \emph{hashdata} layer; and on the bottom a PMA structure for the original
key values, compressed using an augmented version of forward compression known
as the \emph{keydata} layer.

The centroid tree on a high level behaves similarly for static and dynamic
COSB-trees. The leaves of the centroid tree point into the hashdata layer,
where a sequential scan is performed to find the correct key by hash value.
The hashdata layer points directly into the keydata layer, where the
appropriate key is decoded, compared and returned.

@@

\caption{A centroid tree example}
\label{fig:centroid}
\end{figure}

In this approach to solving the dictionary matching problem *cite dictionary
matching problem paper*, a dictionary $D$ of $N$ keys is preprocessed using
hashing and divide-and-conquer via the construction of a centroid tree over
the keys. This technique forms the basis for the Cache Oblivious String B-tree
(COSB-tree) design. Consider the compacted trie $T$ of $D$. Bender et al.
observe that there exists a centroid vertex $\rho$ in $T$ such that $\rho$ has
between N/3 and 2N/3 descendants, inclusive. When searching for a key $k$ in
the compacted trie, you traverse the tree as follows: for each node $y$ that
you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on into the
trie rooted at $y$ or \emph{down trie} of $y$. Otherwise traverse the trie
resulting from excluding $y$ and its subtree. The \emph{centroid tree} is
obtained by rooting the tree at \emph{centroid vertex} $\rho$, and making its
children recursively defined centroid trees of its up and down tries.

At each step through the centroid tree, a consistent fraction of the nodes are
eliminated from the search, resulting in an $O(logN)$ time bound when ignoring
the time required to operate on keys. To make key comparisons faster, the keys
are hashed and hash values are compared in place of the original key values.
To avoid mismatches due to overlap in hash values, the original key values are
compared after finding a match in hash values. The resulting time bound for
\Search{} (interchangeable with \Member{}) is $O(||k||+\log{}N)$. If each node
in the centroid tree knows its successor and predecessor descendants, \Pred{}
and \Succ{} can be achieved in $O(||k||+||k'||+logN)$, where $k'$ is the
successor or predecessor of $k$. Note that the additive $logN$ incurred by the
use of centroid trees can be avoided when using main memory. Centroid trees
are used to capitalize on locality for the cache oblivious string B-tree
design, described later in this section.

The centroid tree is built using $O(N/\log{}N)$ keys, and the centroid tree is
layed out in memory using a modified van Emde Boas (vEB) layout for
weight-balanced binary trees. The \emph{hyperhyperfloors} of nodes are used as
a guide for cutting the tree into equal-size subtrees for the vEB layout. The
\emph{hyperhyperfloor} of x is denoted as $\lhyperfloor{}x\rhyperfloor{}$, and
refers to rounding x down to the nearest power of a power of 2. This ensures
trees of equal size on all \emph{levels of detail}, as the centroid tree is
traversed.

The following approach is used to rebalance the centroid tree: Put all the
elements in DFS or Euler tour order for the trie. Given this order, scan the
trie to identify the centroid vertex, partition the trie into its respective
up and down tries, and recurse on the up and down tries. The Euler tour order
can be achieved via $O(\log{}N)$ scans, since the trie is only $O(\log{}N)$
deep. Thus updates have amortized cost of $O(1+\log^{2}N/B)$ memory transfers.
Note that this doesn't include updating successor or predecessor pointers,
which is described in more detail below.

The following approach is used for maintaining the vEB layout for centroid
trees in the dynamic COSB-tree: first, weight balance is relaxed from a
maximum weight difference of 2 between left and right subtrees, to a larger
constant $c$. Thus a node $v$ only becomes out of balance every
$\Omega(\textsc{weight(v)})$ insertions or deletions of descendants of v,
reducing the frequency of rebuilds. The nodes are stored in a PMA structure,
preserving order of the nodes using $\Theta(N)$space, apart from inserts and
deletes. On insert or delete, the structure is scanned in both directions
starting from the update point for sections of the array to rearrange based on
a "density" threshold of the section. Rearranging here refers to evenly
distributing the nodes across the section to make room for more nodes. To
ensure that updating pointers isn't too expensive, any node included in the
section must have its subtrees included. Specifically, the section is built up
from the the lowest subtree level that contains or contained the update point,
and neighboring subtrees are continuously added until either the density
threshold is met or the entire current subtree layout at this "level of
detail" is included. For example, suppose a node is inserted at the leaf of
some lower recursive subtree $L_{i}$. To continue building up this section,
$L_{i-1}$ or $L_{i+1}$ could be added. If all the lower recursive subtrees
$L_{1}$,$L_{2}$,...,$L_{x}$ have been added, the top level tree $U$ is added.
This corresponds to a subtree $L'_{j}$ one level above. Thus repeat this
process until the density threshold is met, and then rebalance the tree. This
approach depends on the fact that all the recursive subtrees at the same level
of detail asymptotically have the same size, which Bender et al. prove about
the hyperhyperfloor-based vEB layout described previously. This approach
results in an amortized update cost of $O(1+(\log{}N)/B)$ memory transfers per
update. In conjunction with the cost of rebuilds described above, this is only
an additive $O(\log{}N/B)$ factor, so the overall bound is amortized
$O(1+\log^{2}N/B)$ memory transfers.

Note that making the COSB-trees dynamic causes updates to successor and
predecessor pointers in the centroid tree to be expensive. Thus the notion of
ancestors is relaxed to the following (Lemma 10 in \cite{BenderFaKu06}): Let
$\alpha$ be the parent node of $\gamma$ in the compressed trie. Then either
$\alpha$ is a descendant of $\gamma$ or $\gamma$ is a descendant of $\alpha$.
The proof is as follows: As you recursively traverse the trie, both $\alpha$
and $\gamma$ are in the up trie or down trie, until at some point one of them
must be the centroid vertex. This is used to redefine what the predecessor
successor pointers refer to. The centroid may no loner point to the correct
predecessor or successor with respect to the relevant subtrie, but one of the
matched ancestors by construction must point to the correct predecessor or
successor. Thus for the dynamic case, when dealing with successors and
predecessors, $O(logN)$ ancestors' successor and predecessor pointers in
addition to the successor and predecessor pointers of the node in question
must be checked. However, this is an additive factor, so the same time bound
for update described above is still achieved.

\subsubsection{Modifying Front Compression to Preserve Locality}

This section provides a brief summary of how front compression is modified to
optimize compression and decoding for static and dynamic cache-oblivious
string B-trees. Consider a sequence of $j$ keys to store $k_{1}$, $k_{2}$,
..., $k_{j}$. Without compression, this requires $\sum_{j}||k_{j}||$ space in
memory. To compress these keys using front compression, let $\pi_{j}$ be the
longest common prefix between $k_{j}$ and $k_{j+1}$. Now instead we store the
keys in the following format:

@@

The goal of modifying front compression is to maintain some of the space
benefits to compression while reducing the high decoding cost and keeping the
size of D small. Bender et al. provide a modification for static COSB-trees
that reduces the high decoding cost of front compression of D to
$O(1+||k||B)$, with size $O(\llangle{D}\rrangle{})$ after compression. The
modification works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be
the first key added to the compression stream. For each subsequent key
$k_{i}$, if $k_{i}$ can be decoded with just the last $c||k_{i}||$ characters,
add $\pi_{i}$,$\sigma_{i}$ like before. Otherwise, add 0,$k_{i}$, which means
add $k_{i}$ to the stream in its uncompressed form. By construction, this
scheme guarantees that any key $k_{j}$ can be decoded in $O(1+||k||B)$ memory
transfers.%todo: add something about why this isO(<<D>>)?

However, the above modification falls short for anticipating insertions and
deletions. The previous modification relies on each key being decoded within
$c||k^{*}||$ elements of the compressed array, but this is no longer
guaranteed when new keys can be inserted to the left of $k$. To make the
locality-preserving front compression technique suitable for dynamic
COSB-trees, Bender et al. modify the algorithm to include an extra copied
prefix for each key. Consider the $3c||k||$ characters that would be to the
left of a new key$k^{*}$ after insertion. Let $k'$ be the key at the end of
this range. The idea is to prevent the insertion of $k^{*}$ from affecting any
non-copied keys just to the right of $k'$, since this could potentially
increase their decode cost. Compute the length $l$ of the largest common
prefix (lcp) between $k^{*}$ and $k'$, and compare $l$ to all the other lcp's
in this range. $l$ represents the minimum length for any lcp between $k^{*}$
and $k'$, since the keys are stored in lexicographic order. If there is a
copied prefix of length at least $l$ or a copied key within this range, insert
$k^{*}$ as usual. Inserting $k^{*}$ will have no negative effect in this
respect, since the decode length is effectively reset by the copied key or
prefix like before. However, if inserting $k^{*}$ causes the minimum length of
the lcp's in this range to increase, we need to fix the compression. To do
this, we update the lcp of the key directly to the right of $k^{*}$ and store
an extra copy of this lcp with $k^{*}$ itself. In this way, we can charge the
work of copying extra prefixes for a new key to the appropriate number of
elements before it. This number is determined by the length of the extra
copied prefix stored with the key.

@@

The operations supported are dictated by the Dictionary Matching problem.
These operations are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{},
and \Range{}. The behavior of each operation for dynamic COSB-trees is
described below. \Insertkonly{} and \Delete{} are also supported but not
described here.

@@

\Range{}: The paper doesn't say, but presumably this is very similar to
\Search{}. Let $k_{b}$ be the beginning of the range and let $k_{e}$ be the
end of the range. Hash the range keys and identify the predecessor of
$\rho_{b}$ and the successor of $\rho_{e}$ using the centroid tree. Scan the
hashdata layer across from $\rho_{b,p}$ to $\rho_{e,s}$. Only start noting the
beginning of range when the first key with hash value after $\rho_{b}$, and
stop once we reach the first key with hash value after $\rho_{e}$. Decode and
return the set of keys within the given range in the keydata layer.

\subsection{Complexity}

\Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high
probability (w.h.p.). Traversing the centroid tree requires $O(1+\log_{B}N)$
memory transfers, since the tree is built using$O(N/logN)$ keys and is
balanced. Once in the hashdata layer, we are already within $O(logN)$ keys of
$k$. Thus the local scan requires $O(1+(||k||+\log{}N)/B)$ memory transfers.
Obtaining $k$ from the keydata layer requires no more memory transfers than
the scan in the hashdata layer. The \Pred{} and \Succ{} operations are almost
identical in behavior to searches, and thus only have the added cost of having
to operate on $k'$, where $k'$ is the predecessor or successor, respectively.
These operations as a result take $O(1+\log_{B}N+||k||/B+||k'||/B)$ memory
transfers w.h.p.. Similarly, range queries take
$O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory transfers,
where in this case $k'$ is the other end of the range and $Q$ is the set of
keys within the given range.

>>>>

\usepackage{paralist}
\usepackage{graphicx}
\usepackage{url}

@@

\subsection{Description}
Bender et al. perform 2 experiments in \cite{BenderFaKu06}, in which a regular B-tree implementation tuned to various block sizes is compared to a static COSB-tree implementation. The experiment involved measuring the amount of time required to perform: 440,000 and 450,000 inserts of random values, a range query of all values, and 1000 random searches. The results of the experiment show that COSB-trees
outperformed B-trees most of the time for random searches, range queries and
random inserts. However, some point between 440,000 and 450,000 inserts the
COSB-tree had to rebalance the entire tree, causing it to fall behind in this
measurement. However, Bender et al. argue that the experiment was biased
towards the B-tree implementation, since the trees are "young", and thus still
well organized. The COSB-tree implementation is also compared to Berkeley DB
\cite{BerkeleyDB}, and achieves performance comparable to Berkeley DB with
tuned parameters. The results seem to show that there is a window in which
large and small-block size B-trees perform poorly, and that COSB-trees may be
a reasonable alternative to improve performance.

@@

The Dynamic COSB-tree consists of 3 major components: a centroid tree at the
top; a PMA structure of hash data for the keys known as
the \emph{hashdata} layer; and on the bottom a PMA structure for the original
key values, compressed using an augmented version of forward compression known as the \emph{keydata} layer. Each leaf of the centroid tree contains a pointer into the hashdata array. The centroid tree only contains a subset of the keys, so each leaf points to the respective \emph{neighborhood} of the key being searched for. After traversing the centroid tree and following the pointer into the hashdata array layer, we are within $O(\log{}N)$ keys of $k$. Then a sequential scan is performed in the hashdata array layer to find $k$. Each entry in the hashdata array contains the hash value of the corresponding key, and a pointer to the compressed key value in the keydata layer.

@@

\caption{A centroid tree example. Dotted lines represent centroid-rooted up tries, where the parent is not a prefix of the child. Solid lines represent centroid-rooted down tries, where the parent is a prefix of the child.}
\label{fig:centroid}
\end{figure}

COSB-trees are a solution to the dictionary matching problem. Dictionary matching, or string matching, refers to matching strings using prefix search. Let $D$ be a dictionary of $N$ keys. To perform faster searches, the keys in $D$ are preprocessed using hashing and divide-and-conquer via centroid trees. Figure \ref{fig:centroid} depicts an example centroid tree.

Consider the compacted\footnote{where compacted means all non-branching paths are reduced to edges.} trie $T$ constructed over $D$ for string matching.
Consider the vertex $\rho$ in $T$ such that $\rho$ has
between N/3 and 2N/3 descendants, inclusive. Call $\rho$ the \emph{centroid vertex} of $T$. When searching for a key $k$ in
the compacted trie, you traverse the trie as follows: for each node $y$ that
you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on into the
trie rooted at $y$ or \emph{down trie} of $y$. Otherwise traverse the trie
resulting from excluding $y$ and its subtree. The \emph{centroid tree} is
obtained by rooting the tree at the centroid vertex $\rho$, and making its
children recursively defined centroid trees of its up and down tries.

The centroid tree in the COSB-tree is layed out in memory using a modified van Emde Boas (vEB) layout for
weight-balanced binary trees. The \emph{hyperhyperfloors} of nodes are used as
a guide for cutting the tree into equal-size subtrees for the vEB layout. The
\emph{hyperhyperfloor} of x is denoted as $\lhyperfloor{}x\rhyperfloor{}$, and
refers to rounding x down to the nearest power of a power of 2. This ensures
trees of equal size on all \emph{levels of detail}, as the centroid tree is
traversed.

The following approach is used to rebalance the centroid tree: Put all the
elements in DFS or Euler tour order for the compacted trie. Given this order, scan the trie to identify the centroid vertex. Then partition the trie into its respective
up and down tries, and recurse on the up and down tries. The Euler tour order
can be found in $O(\log{}N)$ scans, since the trie is only $O(\log{}N)$
deep. Thus these updates have amortized cost of $O(1+\log^{2}N/B)$ memory transfers.

The following approach is used for maintaining the vEB layout of the centroid
tree in the COSB-tree: first, weight balance is relaxed from a
maximum weight difference of 2 between left and right subtrees, to a larger
constant $c$. Thus a node $v$ only becomes out of balance every
$\Omega(\textsc{weight(v)})$ insertions or deletions of descendants of v. The nodes are stored in a PMA, preserving order of the nodes using $\Theta(N)$space.

On insert or delete, the area around the update point is scanned for sections of the array to rearrange. If a section around the update point reaches a certain "density" threshold, the nodes are evenly
distributed across the section to make room for more nodes. To
ensure that updating pointers isn't too expensive, any node included in the
section must have all of its subtrees included. The subtrees are dictated by the vEB layout, and subtrees are chosen from the bottom up. The smallest neighboring trees (single nodes) in the vEB layout are added first, and larger trees are included when all of their subtrees have been added to the section. Root trees are only added after all bottom trees are added (in terms of the vEB layout). Figure \ref{fig:centroid_relayout} gives an example of how subtrees are chosen for rearranging.

This approach
results in an amortized update cost of $O(1+(\log{}N)/B)$ memory transfers per
update. In conjunction with the cost of rebuilds described above, this is only
an additive $O(\log{}N/B)$ factor, so the overall bound is amortized
$O(1+\log^{2}N/B)$ memory transfers.
\begin{figure}

\begin{center}
	\includegraphics[width=0.8\columnwidth]{figures/centroidrelayoutexample.pdf}
\end{center}

\caption{A centroid layout example. Assume the smallest subtrees in this example only contain a single node. Suppose a node was inserted into a leaf position in the bottom left of the tree. The numbers represent the order in which each subtree is added to the section to be rearranged.}
\label{fig:centroid_relayout}
\end{figure}
\subsubsection{Modifying Front Compression to Preserve Locality}

This section provides a brief summary of how front compression is modified to
optimize compression and decoding for dynamic COSB-trees. Consider a sequence of $j$ keys to store $k_{1}$, $k_{2}$,
..., $k_{j}$. Without compression, this requires $\sum_{j}||k_{j}||$ space in
memory. To compress these keys using standard front compression, let $\pi_{j}$ be the
longest common prefix between $k_{j}$ and $k_{j+1}$. Now instead we store the
keys in the following format:

@@

Bender et al. provide a modification that reduces the high decoding cost of front compressing D to $O(1+||k||B)$, while maintaining a compressed size of $O(\llangle{D}\rrangle{})$\footnote{$\llangle{D}\rrangle{}$ is shorthand for the size of $D$ after front compression} after compression. The modification works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be
the first key added to the compression stream. For each subsequent key
$k_{i}$, if $k_{i}$ can be decoded with just the last $c||k_{i}||$ characters,
add $\pi_{i}$,$\sigma_{i}$ as described above. Otherwise, add $k_{i}$ to the stream without compressing it. By construction, this scheme guarantees that any key $k_{j}$ can be decoded in $O(1+||k||B)$ memory transfers.

However, the above modification falls short for anticipating insertions and
deletions. The previous modification relies on each key being decoded within
$c||k||$ elements of the compressed array, but this is no longer
guaranteed when new keys can be inserted to the left of $k$. To make the
locality-preserving front compression technique suitable for dynamic
COSB-trees, Bender et al. modify the algorithm to include extra copied
prefixes. Suppose a key $k$ is being inserted. If keys further down in the array will be negatively impacted, the longest common prefix between $k$ and the key right after $k$ is copied to $k$ along with $k$'s original compression data. The idea behind adding extra copied prefixes is to maintain $O(||k||)$ decode distances around the area where the new key is being inserted. The characters preceding $k$ are charged for the copied prefix, accounting for the work to add the copied prefix to $k$.

@@

The operations supported by the COSB-tree are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{},
and \Range{}. The behavior of each operation for dynamic COSB-trees is
described below. \Insertkonly{} and \Delete{} are also supported but not
described here.

@@

\Range{}: This operation is not explicitly described in \cite{BenderFaKu06}, but presumably this is very similar to
\Search{}. Let $k_{b}$ be the beginning of the range and let $k_{e}$ be the
end of the range. Hash the range keys and identify the predecessor of
$\rho_{b}$ and the successor of $\rho_{e}$ using the centroid tree. Scan the
hashdata layer across from $\rho_{b,p}$ to $\rho_{e,s}$. Only start noting the
beginning of range when the first key with hash value after $\rho_{b}$, and
stop once we reach the first key with hash value after $\rho_{e}$. Decode and
return the set of keys within the given range in the keydata layer.

\subsection{Complexity}
The keys in are hashed and hash values are compared in place of the original key values to speed up comparisons in the centroid tree and hashdata layer. To avoid mismatches due to overlapping hash values, the original key values are compared only once at the bottom layer, resulting an additive $O(||k||/B)$ memory transfers in each of the operators for each key $k$ that needs to be compared.

\Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high
probability (w.h.p.).  The centroid tree is stored in a PMA with a dynamic CO layout. This layout reduces the number of memory transfers required when traversing the tree to $O(log_{B}N)$.  Once in the hashdata layer, we are already within $O(logN)$ keys of $k$, which require $O(\log{}N)/B)$ memory transfers to scan. Decoding and comparing the original key values requires $O(1+||k||/B)$ memory transfers, due to the modified front compression scheme.

The \Pred{} and \Succ{} operations are similar in behavior to searches, and thus only have the added cost of having to operate on $k'$, where $k'$ is the predecessor or successor, respectively. These operations as a result take $O(1+\log_{B}N+||k||/B+||k'||/B)$ memory transfers w.h.p.. Similarly, range queries take
$O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory transfers,
where in this case $k'$ is the other end of the range and $Q$ is the set of
keys within the given range.
